\documentclass[12pt, leqno]{article}
\input{common}

\begin{document} \hdr{PS 5}{Mon, Mar 2}

{\bf Note:} My solutions this week fit on the front and back of a page
(typed), including the graph for problem 1 and problem text.  Don't
overthink these!

\paragraph*{1: By the book}
Book problem 6.4.1.

\paragraph*{2: QR to SVD}
Suppose $A = QR$ is an economy QR factorization.  Show that the
singular values of $A$ are the same as those of $R$.

\paragraph*{3: Vector projector}
Suppose $A \in \bbR^{m \times n}$ where $m > n$ has full column rank.
Given $A$ and a vector $b$, write one line of MATLAB to compute the
element $c$ in the range space of $A$ that is nearest to $b$
(in the Euclidean norm).

\paragraph*{4: Generally speaking}
Often, we use least squares to construct models of the world.  We
assume that the ``truth'' is
\[
  Ax = b,
\]
but what we measure is the first few rows of $A$ and $b$ (which we
write as $A_1$ and $b_1$), and those measurements are corrupted by
noise.  Suppose we have $A$ exactly, but only get the noisy
partial right hand side $\hat{b}_1 = b_1+e_1$, from which we form
\[
  \mbox{minimize } \|A_1 \hat{x} - \hat{b}_1\|^2.
\]
Our goal in this problem is to use the error analysis ideas in
Section 6.2 to figure out the inherited error in the reconstruction
of $\hat{b}_2 = A_2 \hat{x}$.
\begin{enumerate}
\item
  Let $e_2 = \hat{b}_2-b_2$.  Argue {\em briefly} that
  $e_2 = A_2 A_1^\dagger e_1$.
\item
  Show that
  \[
    \frac{\|e_2\|}{\|b_2\|} \leq \kappa(A_2 A_1^\dagger) \frac{\|e_1\|}{\|b_1\|}.
  \]
\end{enumerate}
Things get somewhat more complicated if we also allow the entries of
$A$ to be contaminated by error, though the same basic ingredients
come into play.

\end{document}
